### 2023-09-06 proofs/tex/constraint-merge-lemmas.tex
Add deterinicity lemmas, get rid of (??)

@@

\begin{lemma} [Completeness of Constraint Entry Merge]
    \label{lemma:entry-merge-completeness}
    For a fixed context $[[Γ]]$,
    suppose that $[[Γ ⊢ scE1]]$ and $[[Γ ⊢ scE2]]$ are matching constraint entries.
    \begin{itemize}
        \item for a type $[[iP]]$ such that $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$,
        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iP : scE]]$.
        \item for a type $[[iN]]$ such that $[[Γ ⊢ iN : scE1]]$ and $[[Γ ⊢ iN : scE2]]$,
        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iN : scE]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us consider the shape of $[[scE1]]$ and $[[scE2]]$.
    \begin{caseof}
        \item $[[scE1]]$ is $[[pua :≈ iQ1]]$ and $[[scE2]]$ is $[[pua :≈ iQ2]]$.
            Then $[[Γ ⊢ iP : scE1]]$ means $[[Γ ⊢ iP ≈ iQ1]]$, 
            and $[[Γ ⊢ iP : scE2]]$ means $[[Γ ⊢ iP ≈ iQ2]]$.
            Then by transitivity of equivalence (\cref{corollary:equivalence-transitivity}),
            $[[Γ ⊢ iQ1 ≈ iQ2]]$, which means $[[nf(iQ1) = nf(iQ2)]]$ by
            \cref{lemma:subt-equiv-algorithmization}.
            Hence, \ruleref{\ottdruleSCMEPEqEqLabel} applies to infer
            $[[Γ ⊢ scE1 & scE2 = scE2]]$, and $[[Γ ⊢ iP : scE2]]$ holds by assumption.
        \item $[[scE1]]$ is $[[pua :≈ iQ1]]$ and $[[scE2]]$ is $[[pua :≥ iQ2]]$.
            Then $[[Γ ⊢ iP : scE1]]$ means $[[Γ ⊢ iP ≈ iQ1]]$, 
            and $[[Γ ⊢ iP : scE2]]$ means $[[Γ ⊢ iP ≥ iQ2]]$.
            Then by transitivity of subtyping, $[[Γ ⊢ iQ1 ≥ iQ2]]$,
            which means $[[Γ ; · ⊨ uQ1 ≥ iQ2 ⫤ ·]]$ by \cref{lemma:pos-subt-completeness}.
            This way, \ruleref{\ottdruleSCMEEqSupLabel} applies to infer
            $[[Γ ⊢ scE1 & scE2 = scE1]]$, and $[[Γ ⊢ iP : scE1]]$ holds by assumption.
        \item $[[scE1]]$ is $[[pua :≥ iQ1]]$ and $[[scE2]]$ is $[[pua :≥ iQ2]]$.
            Then $[[Γ ⊢ iP : scE1]]$ means $[[Γ ⊢ iP ≥ iQ1]]$, 
            and $[[Γ ⊢ iP : scE2]]$ means $[[Γ ⊢ iP ≥ iQ2]]$.
            By the completeness of the least upper bound (\cref{lemma:lub-completeness}), 
            $[[Γ ⊨ iQ1 ∨ iQ2 = iQ]]$, and $[[Γ ⊢ iP ≥ iQ]]$. 
            This way, \ruleref{\ottdruleSCMESupSupLabel} applies to infer
            $[[Γ ⊢ scE1 & scE2 = (pua :≥ iQ)]]$, 
            and $[[Γ ⊢ iP : (pua :≥ iQ)]]$ holds by \ruleref{\ottdruleSATSCESupLabel}.
        \item The negative cases are proved symmetrically.
    \end{caseof}
\end{proof}

@@

\begin{corollary} [Completeness of Unification Constraint Merge] 
    \label{corollary:merge-completeness-unif}
    Suppose that $[[Θ ⊢ UC1 : Ξ1]]$ and $[[Θ ⊢ UC2 : Ξ2]]$.
    If there exists a substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ such that 
    $[[ Θ ⊢ uσ : UC1 ]]$ and $[[ Θ ⊢ uσ : UC2 ]]$
    then $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined.
\end{corollary}

>>>>

\begin{observation}[Constraint Entry Merge is Deterministic]
    \label{obs:entry-merge-deterministic}
    For fixed $[[Γ]]$, $[[scE1]]$, $[[scE2]]$,
    if $[[Γ ⊢ scE1 & scE2 = scE]]$ and $[[Γ ⊢ scE1 & scE2 = scE']]$ 
    then $[[scE]] = [[scE']]$.
\end{observation}
\begin{proof}
    First, notice that the shape of $[[scE1]]$ and $[[scE2]]$
    uniquely determines the rule applied to infer  
    $[[Γ ⊢ scE1 & scE2 = scE]]$,
    which is consequently, the same rule used to 
    infer $[[Γ ⊢ scE1 & scE2 = scE']]$.
    Second, notice that the premises of each rule are deterministic
    on the input:
    the positive subtyping is deterministic by \cref{obs:pos-subt-deterministic},
    and the least upper bound is deterministic by \cref{obs:lub-deterministic}.
\end{proof}

\begin{observation}[Subgyping Constraint Merge is Deterministic]
    \label{obs:subt-merge-deterministic}
    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$
    If $[[Θ ⊢ SC1 & SC2 = UC]]$ and $[[Θ ⊢ SC1 & SC2 = UC']]$ are defined then 
    $[[SC]] = [[SC']]$.
\end{observation}
\begin{proof}
    The proof is analogous to the proof of \cref{obs:unif-merge-deterministic} 
    but uses \cref{obs:entry-merge-deterministic} to show 
    that the merge of the matching constraint entries is fixed.
\end{proof}

@@

\begin{lemma} [Completeness of Constraint Entry Merge]
    \label{lemma:entry-merge-completeness}
    For a fixed context $[[Γ]]$,
    suppose that $[[Γ ⊢ scE1]]$ and $[[Γ ⊢ scE2]]$ are matching constraint entries.
    \begin{itemize}
        \item for a type $[[iP]]$ such that $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$,
        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iP : scE]]$.
        \item for a type $[[iN]]$ such that $[[Γ ⊢ iN : scE1]]$ and $[[Γ ⊢ iN : scE2]]$,
        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iN : scE]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us consider the shape of $[[scE1]]$ and $[[scE2]]$.
    \begin{caseof}
        \item $[[scE1]]$ is $[[pua :≈ iQ1]]$ and $[[scE2]]$ is $[[pua :≈ iQ2]]$.
            The proof repeats the corresponding case of \cref{lemma:unif-entry-merge-completeness}
        \item $[[scE1]]$ is $[[pua :≈ iQ1]]$ and $[[scE2]]$ is $[[pua :≥ iQ2]]$.
            Then $[[Γ ⊢ iP : scE1]]$ means $[[Γ ⊢ iP ≈ iQ1]]$, 
            and $[[Γ ⊢ iP : scE2]]$ means $[[Γ ⊢ iP ≥ iQ2]]$.
            Then by transitivity of subtyping, $[[Γ ⊢ iQ1 ≥ iQ2]]$,
            which means $[[Γ ; · ⊨ uQ1 ≥ iQ2 ⫤ ·]]$ by \cref{lemma:pos-subt-completeness}.
            This way, \ruleref{\ottdruleSCMEEqSupLabel} applies to infer
            $[[Γ ⊢ scE1 & scE2 = scE1]]$, and $[[Γ ⊢ iP : scE1]]$ holds by assumption.
        \item $[[scE1]]$ is $[[pua :≥ iQ1]]$ and $[[scE2]]$ is $[[pua :≥ iQ2]]$.
            Then $[[Γ ⊢ iP : scE1]]$ means $[[Γ ⊢ iP ≥ iQ1]]$, 
            and $[[Γ ⊢ iP : scE2]]$ means $[[Γ ⊢ iP ≥ iQ2]]$.
            By the completeness of the least upper bound (\cref{lemma:lub-completeness}), 
            $[[Γ ⊨ iQ1 ∨ iQ2 = iQ]]$, and $[[Γ ⊢ iP ≥ iQ]]$. 
            This way, \ruleref{\ottdruleSCMESupSupLabel} applies to infer
            $[[Γ ⊢ scE1 & scE2 = (pua :≥ iQ)]]$, 
            and $[[Γ ⊢ iP : (pua :≥ iQ)]]$ holds by \ruleref{\ottdruleSATSCESupLabel}.
        \item The negative cases are proved symmetrically.
    \end{caseof}
\end{proof}

@@

### 2023-09-06 proofs/tex/singularity-lemmas.tex
Grammar update: Γ = set
Add deterinicity lemmas, get rid of (??)

            By \cref{lemma:shape-supertypes-norm}, 
            the last subtyping means that $[[nf(iP') = ∃nas.↓iN']]$,
            such that
            \begin{enumerate}
                \item $[[Γ, nbj, nas ⊢ iN']]$
                \item $[[ord {nas} in iN' = nas]]$
                \item for some substitution $[[Γ, nbj ⊢ σ : nas]]$, 
                    $[[ [σ]iN' = nbj ]]$.
            \end{enumerate}
            Since $[[nbj ∉ fv(nf(iP'))]]$,
            the latter means that $[[iN' = na]]$, and then 
            $[[nf(iP') = ∃na.↓na]]$ for some $[[na]]$.
            Finally, notice that all the types of shape
            $[[∃na.↓na]]$ are equal.
   \end{caseof}

@@

\begin{lemma}[Soundness of Singularity]
    \label{lemma:singularity-soundness}
    Suppose $[[Θ ⊢ SC : Ξ]]$, and $[[SC singular with uσ]]$. 
    Then $[[ Θ ⊢ uσ : Ξ ]]$,
     $[[ Θ ⊢ uσ : SC ]]$ and for any 
    $[[uσ']]$ such that $[[Θ ⊢ uσ : SC]]$,
    $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$.
\end{lemma}
\begin{proof}
    Suppose that $(Θ  ⊢ SC) ⊢ uσ'$.
    It means that for every $[[scE]] \in [[SC]]$ restricting $[[α̂±]]$,
    $[[Θ(α̂±) ⊢ [uσ']α̂± : scE]]$ holds.
    $[[SC singular with uσ]]$ means $[[scE singular with [uσ]α̂± ]]$,
    and hence, by \cref{lemma:entry-singularity-completeness},
    $[[ Θ(α̂±) ⊢ [uσ']α̂± ≈ [uσ]α̂±  ]]$ holds.

@@

>>>>

            By \cref{lemma:shape-supertypes-norm}, 
            the last subtyping means that $[[nf(iP') = ∃nas.↓iN']]$,
            such that
            \begin{enumerate}
                \item $[[Γ, nbj, nas ⊢ iN']]$
                \item $[[ord {nas} in iN' = nas]]$
                \item for some substitution $[[Γ, nbj ⊢ σ :{nas}]]$, 
                    $[[ [σ]iN' = nbj ]]$.
            \end{enumerate}
            Since $[[nbj ∉ fv(nf(iP'))]]$,
            the latter means that $[[iN' = na]]$, and then 
            $[[nf(iP') = ∃na.↓na]]$ for some $[[na]]$.
            Finally, notice that all the types of shape
            $[[∃na.↓na]]$ are equal.
   \end{caseof}

@@

\begin{lemma}[Soundness of Singularity]
    \label{lemma:singularity-soundness}
    Suppose $[[Θ ⊢ SC : Ξ]]$, and $[[SC singular with uσ]]$. 
    Then $[[ Θ ⊢ uσ : Ξ ]]$,
     $[[ Θ ⊢ uσ : SC ]]$ and for any 
    $[[uσ']]$ such that $[[Θ ⊢ uσ : SC]]$,
    $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$.
\end{lemma}
\begin{proof}
    Suppose that $[[Θ ⊢ uσ' : SC]]$.
    It means that for every $[[scE]] \in [[SC]]$ restricting $[[α̂±]]$,
    $[[Θ(α̂±) ⊢ [uσ']α̂± : scE]]$ holds.
    $[[SC singular with uσ]]$ means $[[scE singular with [uσ]α̂± ]]$,
    and hence, by \cref{lemma:entry-singularity-completeness},
    $[[ Θ(α̂±) ⊢ [uσ']α̂± ≈ [uσ]α̂±  ]]$ holds.

@@

\begin{observation}[Singularity is Deterministiic]
    For a fixed $[[SC]]$ such that $[[Θ ⊢ SC : Ξ]]$,
    if $[[SC singular with uσ]]$ and $[[SC singular with uσ']]$,
    then $[[uσ = uσ']]$.
\end{observation}
\begin{proof}
    By \label{lemma:singularity-soundness},
    $[[Θ ⊢ uσ : Ξ]]$ and $[[Θ ⊢ uσ' : Ξ]]$.
    It means that both $[[uσ]]$ and $[[uσ']]$
    act as identity outside of $[[Ξ]]$.

    Moreover, for any $[[α̂± ∊ Ξ]]$,
    $[[Θ ⊢ SC : Ξ]]$ means that 
    there is a unique $[[scE ∊ SC]]$ restricting $[[α̂±]]$.
    Then $[[SC singular with uσ]]$ means 
    that $[[scE singular with [uσ]α̂± ]]$.
    By looking at the inference rules, it is easy to see that
    $[[ [uσ]α̂± ]]$ is uniquely determined by $[[scE]]$, which, 
    Similarly, $[[ [uσ']α̂± ]]$ is also uniquely determined by $[[scE]]$,  
    in the same way, and hence, $[[ [uσ]α̂± = [uσ']α̂± ]]$.
\end{proof}

### 2023-09-06 proofs/tex/unif-constraint-merge-lemmas.tex
Add deterinicity lemmas, get rid of (??)

@@

        If $[[ucE]]$ is from the first or the second part, 
        then $[[ucE]] \in [[UC1]] \cup [[UC2]]$ holds immediately.
        If $[[ucE]]$ is from the third part,
        then $[[ucE]]$ is the merge of two matching entries
        $[[ucE1]] \in [[UC1]]$ and $[[ucE2]] \in [[UC2]]$.
        Since $[[UC1]]$ and $[[UC2]]$ are normalized unification , 
        $[[ucE1]]$ and $[[ucE2]]$ have one of the following forms:
        \begin{itemize}
            \item $[[α̂⁺ :≈ iP1]]$ and $[[α̂⁺ :≈ iP2]]$, 
                where $[[iP1]]$ and $[[iP2]]$ are normalized,
                and then since $[[Θ(α̂⁺) ⊢ ucE1 & ucE2 = ucE]]$ exists, 
                \ruleref{\ottdruleUCMEPEqEqLabel} was applied to infer it.
                It means that $[[ucE]] = [[ucE1]] = [[ucE2]]$;
            \item $[[α̂⁻ :≈ iN1]]$ and $[[α̂⁻ :≈ iN2]]$, 
               then symmetrically, 
               $[[Θ(α̂⁻) ⊢ ucE1 & ucE2 = ucE]] = [[ucE1]] = [[ucE2]]$
        \end{itemize}
        In both cases, $[[ucE]] \in [[UC1]] \cup [[UC2]]$.

        \item $[[UC1]] \cup [[UC2]] \subseteq [[UC1 & UC2]]$\\
        Let us take 
        an arbitrary $[[ucE1]] \in [[UC1]]$.
        Then since $[[UC1]]$ is a unification constraint,
         $[[ucE1]]$ has one of the following forms:
        \begin{itemize}
            \item $[[α̂⁺ :≈ iP]]$ where $[[iP]]$ is normalized.
            If $[[α̂⁺]] \notin [[dom(UC2)]]$, then $[[ucE1]] \in [[UC1 & UC2]]$.
            Otherwise, there is a normalized matching
            $[[ucE2]] = [[(α̂⁺ :≈ iP')]] \in [[UC2]]$ and then
            since $[[UC1 & UC2]]$ exists, 
            \ruleref{\ottdruleUCMEPEqEqLabel} was applied to construct
            $[[ucE1 & ucE2]] \in [[UC1 & UC2]]$.
            By inversion of \ruleref{\ottdruleUCMEPEqEqLabel},
            $[[ucE1 & ucE2]] = [[ucE1]]$, and
            $[[nf(iP) = nf(iP')]]$, which since $[[iP]]$
            and $[[iP']]$ are normalized, implies that $[[iP = iP']]$, 
            that is $[[ucE1]] = [[ucE2]] \in [[UC1 & UC2]]$.
            \item $[[α̂⁻ :≈ iN]]$ where $[[iN]]$ is normalized.
            Then symmetrically, $[[ucE1]] = [[ucE2]] \in [[UC1 & UC2]]$.
        \end{itemize}
        Similarly, if we take an arbitrary $[[ucE2]] \in [[UC2]]$,
        then $[[ucE1]] = [[ucE2]] \in [[UC1 & UC2]]$. 
    \end{itemize}
\end{proof}

@@

\begin{lemma} [Completeness of Unification Constraint Entry Merge]
    \label{lemma:unif-entry-merge-completeness}
    For a fixed context $[[Γ]]$,
    suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$ are matching constraint entries.
    \begin{itemize}
        \item for a type $[[iP]]$ such that $[[Γ ⊢ iP : ucE1]]$ and $[[Γ ⊢ iP : ucE2]]$,
        $[[Γ ⊢ ucE1 & ucE2 = ucE]]$ is defined and $[[Γ ⊢ iP : ucE]]$.
        \item for a type $[[iN]]$ such that $[[Γ ⊢ iN : ucE1]]$ and $[[Γ ⊢ iN : ucE2]]$,
        $[[Γ ⊢ ucE1 & ucE2 = ucE]]$ is defined and $[[Γ ⊢ iN : ucE]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    The proof repeats the one of \cref{lemma:entry-merge-completeness}
    and is done by the case analysis on the shape of $[[ucE1]]$ and $[[ucE2]]$.
    However, it only needs to consider two cases.
    \begin{caseof}
        \item $[[ucE1]]$ is $[[pua :≈ iQ1]]$ and $[[ucE2]]$ is $[[pua :≈ iQ2]]$.
        \item $[[ucE1]]$ is $[[nua :≈ iN1]]$ and $[[ucE2]]$ is $[[nua :≈ iM2]]$.
    \end{caseof}
    The proof of these cases is based only on \cref{lemma:subt-equiv-algorithmization}
    and \cref{corollary:equivalence-transitivity}, and does not require the properties of 
    the least upper bound or subtyping.
\end{proof}

\begin{lemma} [Completeness of Unification Constraint Merge] 
    \label{lemma:unif-merge-completeness}
    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$.
    Then for any $[[Ξ]] \supseteq [[dom(UC1) ∪ dom(UC2)]]$ 
    and substitution $[[Θ ⊢ uσ : Ξ]]$ such that $[[ Θ ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$, 
    \begin{enumerate}
        \item $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and
        \item $[[ Θ   ⊢ uσ : UC ]]$.
    \end{enumerate}
\end{lemma}
\begin{proof}
    The proof repeats the proof of \cref{lemma:merge-completeness} but 
    uses \cref{lemma:unif-entry-merge-completeness} instead of \cref{lemma:entry-merge-completeness}.
\end{proof}

>>>>

\begin{observation} [Unification Constraint Merge is Deterministic]
    \label{obs:unif-merge-deterministic}
    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$
    If $[[Θ ⊢ UC1 & UC2 = UC]]$ and $[[Θ ⊢ UC1 & UC2 = UC']]$ are defined then 
    $[[UC]] = [[UC']]$.
\end{observation}
\begin{proof}
    $[[UC]]$ and $[[UC']]$ both consists of three parts: 
    Entries of $[[UC1]]$ that do not have matching entries in $[[UC2]]$,
    entries of $[[UC2]]$ that do not have matching entries in $[[UC1]]$,
    and the merge of matching entries.

    The parts corresponding to unmatched entries of $[[UC1]]$ and $[[UC2]]$ coincide, 
    since $[[UC1]]$ and $[[UC2]]$ are fixed.
    To show that the merge of matching entries coincide,
    let us take any pair of matching $[[ucE1 ∊ UC1]]$ and $[[ucE2 ∊ UC2]]$
    and consider their shape.
    \begin{caseof}
        \item $[[ucE1]]$ is $[[pua :≈ iQ1]]$ and $[[ucE2]]$ is $[[pua :≈ iQ2]]$
            then the result, if it exists, is always $[[ucE1]]$,
            by inversion of \ruleref{\ottdruleSCMEPEqEqLabel}.
        \item $[[ucE1]]$ is $[[nua :≈ iN1]]$ and $[[ucE2]]$ is $[[nua :≈ iN2]]$
            then analogously, the result, if it exists, is always $[[ucE1]]$,
            by inversion of \ruleref{\ottdruleSCMENEqEqLabel}.
    \end{caseof}
    This way, the third group of entries coincide as well.
\end{proof}

@@

        If $[[ucE]]$ is from the first or the second part, 
        then $[[ucE]] \in [[UC1]] \cup [[UC2]]$ holds immediately.
        If $[[ucE]]$ is from the third part,
        then $[[ucE]]$ is the merge of two matching entries
        $[[ucE1]] \in [[UC1]]$ and $[[ucE2]] \in [[UC2]]$.
        Since $[[UC1]]$ and $[[UC2]]$ are normalized unification , 
        $[[ucE1]]$ and $[[ucE2]]$ have one of the following forms:
        \begin{itemize}
            \item $[[α̂⁺ :≈ iP1]]$ and $[[α̂⁺ :≈ iP2]]$, 
                where $[[iP1]]$ and $[[iP2]]$ are normalized,
                and then since $[[Θ(α̂⁺) ⊢ ucE1 & ucE2 = ucE]]$ exists, 
                \ruleref{\ottdruleSCMEPEqEqLabel} was applied to infer it.
                It means that $[[ucE]] = [[ucE1]] = [[ucE2]]$;
            \item $[[α̂⁻ :≈ iN1]]$ and $[[α̂⁻ :≈ iN2]]$, 
               then symmetrically, 
               $[[Θ(α̂⁻) ⊢ ucE1 & ucE2 = ucE]] = [[ucE1]] = [[ucE2]]$
        \end{itemize}
        In both cases, $[[ucE]] \in [[UC1]] \cup [[UC2]]$.

        \item $[[UC1]] \cup [[UC2]] \subseteq [[UC1 & UC2]]$\\
        Let us take 
        an arbitrary $[[ucE1]] \in [[UC1]]$.
        Then since $[[UC1]]$ is a unification constraint,
         $[[ucE1]]$ has one of the following forms:
        \begin{itemize}
            \item $[[α̂⁺ :≈ iP]]$ where $[[iP]]$ is normalized.
            If $[[α̂⁺]] \notin [[dom(UC2)]]$, then $[[ucE1]] \in [[UC1 & UC2]]$.
            Otherwise, there is a normalized matching
            $[[ucE2]] = [[(α̂⁺ :≈ iP')]] \in [[UC2]]$ and then
            since $[[UC1 & UC2]]$ exists, 
            \ruleref{\ottdruleSCMEPEqEqLabel} was applied to construct
            $[[ucE1 & ucE2]] \in [[UC1 & UC2]]$.
            By inversion of \ruleref{\ottdruleSCMEPEqEqLabel},
            $[[ucE1 & ucE2]] = [[ucE1]]$, and
            $[[nf(iP) = nf(iP')]]$, which since $[[iP]]$
            and $[[iP']]$ are normalized, implies that $[[iP = iP']]$, 
            that is $[[ucE1]] = [[ucE2]] \in [[UC1 & UC2]]$.
            \item $[[α̂⁻ :≈ iN]]$ where $[[iN]]$ is normalized.
            Then symmetrically, $[[ucE1]] = [[ucE2]] \in [[UC1 & UC2]]$.
        \end{itemize}
        Similarly, if we take an arbitrary $[[ucE2]] \in [[UC2]]$,
        then $[[ucE1]] = [[ucE2]] \in [[UC1 & UC2]]$. 
    \end{itemize}
\end{proof}

@@

\begin{lemma} [Completeness of Unification Constraint Entry Merge]
    \label{lemma:unif-entry-merge-completeness}
    For a fixed context $[[Γ]]$,
    suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$ are matching constraint entries.
    \begin{itemize}
        \item for a type $[[iP]]$ such that $[[Γ ⊢ iP : ucE1]]$ and $[[Γ ⊢ iP : ucE2]]$,
        $[[Γ ⊢ ucE1 & ucE2 = ucE]]$ is defined and $[[Γ ⊢ iP : ucE]]$.
        \item for a type $[[iN]]$ such that $[[Γ ⊢ iN : ucE1]]$ and $[[Γ ⊢ iN : ucE2]]$,
        $[[Γ ⊢ ucE1 & ucE2 = ucE]]$ is defined and $[[Γ ⊢ iN : ucE]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us consider the shape of $[[ucE1]]$ and $[[ucE2]]$.
    \begin{caseof}
        \item $[[ucE1]]$ is $[[pua :≈ iQ1]]$ and $[[ucE2]]$ is $[[pua :≈ iQ2]]$.
            Then $[[Γ ⊢ iP : ucE1]]$ means $[[Γ ⊢ iP ≈ iQ1]]$, 
            and $[[Γ ⊢ iP : ucE2]]$ means $[[Γ ⊢ iP ≈ iQ2]]$.
            Then by transitivity of equivalence (\cref{corollary:equivalence-transitivity}),
            $[[Γ ⊢ iQ1 ≈ iQ2]]$, which means $[[nf(iQ1) = nf(iQ2)]]$ by
            \cref{lemma:subt-equiv-algorithmization}.
            Hence, \ruleref{\ottdruleSCMEPEqEqLabel} applies to infer
            $[[Γ ⊢ ucE1 & ucE2 = ucE2]]$, and $[[Γ ⊢ iP : ucE2]]$ holds by assumption.
        \item $[[ucE1]]$ is $[[nua :≈ iN1]]$ and $[[ucE2]]$ is $[[nua :≈ iM2]]$.
            The proof is symmetric.
    \end{caseof}
\end{proof}

\begin{lemma} [Completeness of Unification Constraint Merge] 
    \label{lemma:unif-merge-completeness}
    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$.
    Then for any $[[Ξ]] \supseteq [[dom(UC1) ∪ dom(UC2)]]$ 
    and substitution $[[Θ ⊢ uσ : Ξ]]$ such that $[[ Θ ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$, 
    \begin{enumerate}
        \item $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and
        \item $[[ Θ   ⊢ uσ : UC ]]$.
    \end{enumerate}
\end{lemma}
\begin{proof}
    The proof repeats the proof of \cref{lemma:merge-completeness}
    for cases 
    uses \cref{lemma:unif-entry-merge-completeness} instead of \cref{lemma:entry-merge-completeness}.
\end{proof}

### 2023-09-06 proofs/tex/unification-lemmas.tex
Grammar update: Γ = set
Add deterinicity lemmas, get rid of (??)

@@

            Let us take an arbitrary $[[Θ ⊢ uσ : α̂⁻ ]]$ such that $[[ [uσ]α̂⁻ = iM ]]$.
            $[[Θ ⊢ uσ : α̂⁻]]$ means that $[[Δ ⊢ iM]]$.
            This way, \ruleref{\ottdruleUNUVarLabel} is applicable to infer 
            $[[Γ ; Θ ⊨ â⁻ ≈u iM ⫤ (â⁻ :≈ iM)]]$.

>>>>

\begin{observation}[Unification is deterministic]
    \label{obs:unif-deterministic}
    \hfill
    \begin{itemize}
        \item [$+$] If $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$
            and $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC']]$ then $[[UC = UC']]$.
        \item [$-$] If $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$
            and $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC']]$ then $[[UC = UC']]$.
    \end{itemize}
\end{observation}
\begin{proof}
    We prove it by mutual structural induction on 
    $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ and $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC']]$.
    Let us consider the positive case, since the negative case is symmetric.

    First, notice that the rule applied the last is uniquely determined by the
    shape of $[[uP]]$ and $[[iQ]]$.
    Second, the premises of each rule are
    deterministic on the input either by the induction 
    hypothesis or by \cref{obs:unif-merge-deterministic}.
\end{proof}

@@

            Let us take an arbitrary $[[Θ ⊢ uσ : {α̂⁻} ]]$ such that $[[ [uσ]α̂⁻ = iM ]]$.
            $[[Θ ⊢ uσ : {α̂⁻}]]$ means that $[[Δ ⊢ iM]]$.
            This way, \ruleref{\ottdruleUNUVarLabel} is applicable to infer 
            $[[Γ ; Θ ⊨ â⁻ ≈u iM ⫤ (â⁻ :≈ iM)]]$.
